Notes on Control Plane Inference

### 2021-02-03 inference.tex
write notes about bug

\usepackage{amsmath, amsthm, array, stmaryrd}
\usepackage{booktabs}
\usepackage{todonotes}

@@

Given a program $p$, compute $\varphi$ such that for all
instantiations $\sigma$, $\sigma \models \varphi$ if (and
only if) one of the following (equivalent expressions) holds:
\begin{enumerate}
\item $\forall \mathit{pkt}. \llbracket p[\sigma]\rrbracket~\mathit{pkt} \neq \bot$
\item $\mathit{wp}(p[\sigma], \top)$ is valid
\end{enumerate}

@@

Now we can reformulate the problem entirely logically. Given a program
$p$ and its corresponding instrumented verification condition
$\psi_p$, compute a controller constraint $\varphi$ such that the following equation is valid
\[\forall \vec{\texttt{d}}.~\psi_p[\vec{\texttt{d}},\vec{\mathsf{c}}] \iff \varphi[\vec{\mathsf{c'}}] \]
Now we only need to investigate a quantifier elimination algorithm for
the variables $\vec{\texttt{d}}$. The standard algorithm (replace
every $\texttt{d}_i$ of sort $\texttt{(\_~BitVec~n)}$ with every
element in $\mathbf{2}^n$) for fixed-width bitvector theories is
intractable: a standard ethernet frame is $1500$ bits, and the number
of corresponding disjunctions is $2^{1500} = 10^{451.5\ldots}$, which
is far greater than the $10^{82}$ atoms in the universe.

@@

\section{An Incremental Problem Statement}

>>>>

\usepackage{amsmath, amsthm, array, stmaryrd, latexsym}
\usepackage{booktabs}
\usepackage{todonotes}
\usepackage{float}

@@

Given a program $p$, compute $\varphi$ such that for all
instantiations $\sigma$, $\sigma \models \varphi$ if (and
only if) one of the following (equivalent expressions) holds:
\begin{enumerate}
\item $\forall \mathit{pkt}. \llbracket p[\sigma]\rrbracket~\mathit{pkt} \neq \mathsf{error}$
\item $\mathit{wp}(p[\sigma], \top)$ is valid
\end{enumerate}

@@

Now we can reformulate the problem entirely logically. Given a program
$p$ and its corresponding instrumented verification condition
$\psi_p$, compute a controller constraint $\varphi$ such that the following equation is valid
%% \[(\forall \vec{\texttt{d}}.~\psi_p[\vec{\texttt{d}},\vec{\mathsf{c}}]) \iff \varphi[\vec{\mathsf{c'}}] \]
\[\varphi[\vec{\mathsf{c}}] \Rightarrow \psi_p[\vec{\texttt{d}},\vec{\mathsf{c}}] \]

Now we only need to investigate a quantifier elimination algorithm for
the variables $\vec{\texttt{d}}$. The standard algorithm (replace
every $\texttt{d}_i$ of sort $\texttt{(\_~BitVec~n)}$ with every
element in $\mathbf{2}^n$) for fixed-width bitvector theories is
intractable: a standard ethernet frame is $1500$ bits, and the number
of corresponding disjunctions is $2^{1500} = 10^{451.5\ldots}$, which
is far greater than the $10^{82}$ atoms in the universe.

@@

\section{Examples}

There are a few questions we want to investigate with examples.
\begin{enumerate}
\item \label{question:rowiter}
  Do we need to consider multiple rows at the same time? In other words, do
  we need to instrument each table $t$ with a different row for \(t[0]\ldots
  t[t.\mathsf{maxrows}]\) or is it sufficient to, like P4v, only consider a
  single row at a time?
\item \label{question:elimination}
  Can we eliminate all references to dataplane state in the conditions we
  generate?
\end{enumerate}

\subsection{A Multi-Table Example}
We will construct a pipeline with 3 tables, \(t_i\) for \(i=0,1,2\),
where each \(t_i\) matches on \(x_i\) and assigns a value to
\(y_i\).
\begin{verbatim}
table ti {
  keys : xi
  actions : ai
  default : ai(i)
  max_rows : 2
}
action ai(y){
  yi := y
}
\end{verbatim}

The control block applies \(t_0\), and then conditionally applies
\(t_1\) or \(t_2\). The assertion is that \(t_i\) is even for all \(i=0,1,2\).

\begin{verbatim}
y0 := 0;
y1 := 0;
y2 := 0;
t0.apply();
if (y0%3 = 0) {
  t1.apply()
} else {
  t2.apply()
}
assert y0%2 = 0;
assert y1%2 = 0;
assert y2%2 = 0;
\end{verbatim}

Now, we can instrument this program using the ghost state defined above.
\begin{verbatim}
y0 := 0; y1 := 0; y2 := 0;
assume t0[0].action <= 0;
assume t0[1].action <= 0;
assume t0.numrows <= 2;
{ assume t0.numrows > 0 && t0[0].keys(x0) = x0;
  assume t0[0].action = 0;
  y0 := t0[0].data(y)
} [] {
 assume not (t0.numrows > 0 && t0[0].keys(x0) = x0);
 assume t0.numrows > 1  && t0[1].keys(x0) = x0;
 assume t0[1].action = 0;
 y0 := t0[1].data(y)
} [] {
 assume not (t0.numrows > 0 && t0[0].keys(x0) = x0);
 assume not (t0.numrows > 1 && t0[1].keys(x0) = x0);
 y0 := 0
};
{assume y0%3 = 0;
  assume t1[0].action <= 0;
  assume t1[1].action <= 0;
  assume t1.numrows <= 2;
  { assume t1.numrows > 0 && t1[0].keys(x1) = x1;
    assume t1[0].action = 0;
    y1 := t1[0].data(y)
  } [] {
    assume not (t1.numrows > 0 && t1[0].keys(x1) = x1);
    assume t1.numrows > 1  && t1[1].keys(x1) = x1;
    assume t1[1].action = 0;
    y1 := t1[1].data(y)
  } [] {
    assume not (t1.numrows > 0 && t1[0].keys(x1) = x1);
    assume not (t1.numrows > 1 && t1[1].keys(x1) = x1);
    y1 := 1
  }
} [] {
  assume y0%3 != 0;
  assume t2[0].action <= 0;
  assume t2[1].action <= 0;
  assume t2.numrows <= 2;
  { assume t2.numrows > 0 && t2[0].keys(x2) = x2;
    assume t2[0].action = 0;
    y2 := t2[0].data(y)
  } [] {
    assume not (t2.numrows > 0 && t2[0].keys(x2) = x2);
    assume t2.numrows > 1  && t2[1].keys(x2) = x2;
    assume t2[1].action = 0;
    y2 := t2[1].data(y)
  } [] {
    assume not (t2.numrows > 0 && t2[0].keys(x2) = x2);
    assume not (t2.numrows > 1 && t2[1].keys(x2) = x2);
    y2 := 2
  }
}
assert y0 % 2 = y1 % 2 = y2 % 2 = 0
\end{verbatim}

And finally, we can compute the weakest precondition to produce a (universally
closed) condition on variables.

\[
\begin{array}{l}
  t_0.\NumRows > 0 \wedge t_0[0].\MatchRow(x_0) = x_0 \\
  \Rightarrow
  \begin{array}[t]{l}
    t_0[0].\ActionDataRow(y) \% 3 = 0 \Rightarrow
    \begin{array}[t]{l}
      t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1 \\
      \qquad \Rightarrow t_1[0].\ActionDataRow(y) \% 2 = 0 \wedge t_0[0].\ActionDataRow(y)\% 2 = 0 \\
      \wedge (\neg (t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1) \\
      \phantom{\wedge(}\wedge (t_1.\NumRows > 1 \wedge t_1[1].\MatchRow(x_1) = x_1) \\
      \phantom{\wedge(}\Rightarrow t_1[1].\ActionDataRow(y) \% 2 = 0 \wedge t_0[1].\ActionDataRow(y)\% 2 = 0 )\\
      \wedge (\neg (t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1) \\
      \phantom{\wedge (}\wedge \neg (t_1.\NumRows > 1 \wedge t_1[1].\MatchRow(x_1) = x_1)) \\
    \end{array} \\
    \wedge\,t_0[0].\ActionDataRow(y) \% 3 \neq 0 \Rightarrow
    \begin{array}[t]{l}
      t_2.\NumRows > 0 \wedge t_2[0].\MatchRow(x_2) = x_2 \\
      \qquad \Rightarrow t_2[0].\ActionDataRow(y) \% 2 = 0 \wedge t_0[0].\ActionDataRow(y)\% 2 = 0 \\
      \wedge (\neg (t_2.\NumRows > 0 \wedge t_2[0].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge(}\wedge (t_2.\NumRows > 1 \wedge t_2[1].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge(}\Rightarrow t_2[1].\ActionDataRow(y) \% 2 = 0 \wedge t_0[1].\ActionDataRow(y)\% 2 = 0 )\\
      \wedge (\neg (t_2.\NumRows > 0 \wedge t_2[0].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge (}\wedge \neg (t_2.\NumRows ? 1 \wedge t_2[1].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge (} t_0[1].\ActionDataRow(y) \% 2 = 0)
    \end{array}
  \end{array} \\
  t_0.\NumRows > 1 \wedge t_0[1].\MatchRow(x_0) = x_0 \\
  \wedge \neg (t_0.\NumRows > 0 \wedge t_0[0].\MatchRow(x_0) = x_0 ) \\
  \Rightarrow
  \begin{array}[t]{l}
    t_0[1].\ActionDataRow(y) \% 3 = 0 \Rightarrow
    \begin{array}[t]{l}
      t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1 \\
      \qquad \Rightarrow t_1[0].\ActionDataRow(y) \% 2 = 0 \wedge t_0[1].\ActionDataRow(y)\% 2 = 0 \\
      \wedge (\neg (t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1) \\
      \phantom{\wedge(}\wedge (t_1.\NumRows > 1 \wedge t_1[1].\MatchRow(x_1) = x_1) \\
      \phantom{\wedge(}\Rightarrow t_1[1].\ActionDataRow(y) \% 2 = 0 \wedge t_0[1].\ActionDataRow(y)\% 2 = 0 )\\
      \wedge ((t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1) \\
      \phantom{\wedge (}\vee (t_1.\NumRows ? 1 \wedge t_1[1].\MatchRow(x_1) = x_1)) \\
    \end{array} \\
    \wedge\,t_0[1].\ActionDataRow(y) \% 3 \neq 0 \Rightarrow
    \begin{array}[t]{l}
      t_2.\NumRows > 0 \wedge t_2[0].\MatchRow(x_2) = x_2 \\
      \qquad \Rightarrow t_2[0].\ActionDataRow(y) \% 2 = 0 \wedge t_0[1].\ActionDataRow(y)\% 2 = 0 \\
      \wedge (\neg (t_2.\NumRows > 0 \wedge t_2[0].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge(}\wedge (t_2.\NumRows > 1 \wedge t_2[1].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge(}\Rightarrow t_2[1].\ActionDataRow(y) \% 2 = 0 \wedge t_0[1].\ActionDataRow(y)\% 2 = 0 )\\
      \wedge (\neg (t_2.\NumRows > 0 \wedge t_2[0].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge (}\wedge \neg (t_2.\NumRows ? 1 \wedge t_2[1].\MatchRow(x_2) = x_2) \\
      \phantom{\wedge (} t_0[1].\ActionDataRow(y) \% 2 = 0)
    \end{array}
  \end{array} \\
  \neg (t_0.\NumRows > 0 \wedge t_0[0].\MatchRow(x_0) = x_0) \\
  \wedge \neg (t_0.\NumRows > 1 \wedge t_0[1].\MatchRow(x_0) = x_0 ) \\
  \Rightarrow
  \begin{array}[t]{l}
    t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1 \\
    \qquad \Rightarrow t_1[0].\ActionDataRow(y) \% 2 = 0 \\
    \wedge (\neg (t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1) \\
    \phantom{\wedge(}\wedge (t_1.\NumRows > 1 \wedge t_1[1].\MatchRow(x_1) = x_1) \\
    \phantom{\wedge(}\Rightarrow t_1[1].\ActionDataRow(y) \% 2 = 0 )\\
    \wedge ((t_1.\NumRows > 0 \wedge t_1[0].\MatchRow(x_1) = x_1) \\
    \phantom{\wedge (}\vee (t_1.\NumRows ? 1 \wedge t_1[1].\MatchRow(x_1) = x_1)) \\
    \end{array}
\end{array}
\]

\subsubsection{Question~\ref{question:rowiter}}

For this example, its clear that we dont need the cases for both \(t_i[0]\) and
\(t_i[1]\) for $i = 1,2,3$, so we can equivalently compute the following
(universally closed) verification condition.

\[
\begin{array}{l}
  t_0.\MatchRow(x_0) = x_0 \\
  \Rightarrow
  \begin{array}[t]{l}
    t_0.\ActionDataRow(y) \% 3 = 0 \Rightarrow
    \begin{array}[t]{l}
      t_1.\MatchRow(x_1) = x_1 \\
      \qquad \Rightarrow t_1.\ActionDataRow(y) \% 2 = 0 \wedge t_0.\ActionDataRow(y)\% 2 = 0 \\
      \wedge \neg (t_1.\MatchRow(x_1) = x_1 \Rightarrow \bot) \\
    \end{array} \\
    \wedge\,t_0.\ActionDataRow(y) \% 3 \neq 0 \Rightarrow
    \begin{array}[t]{l}
      t_2.\MatchRow(x_2) = x_2 \\
      \qquad \Rightarrow t_2.\ActionDataRow(y) \% 2 = 0 \wedge t_0.\ActionDataRow(y)\% 2 = 0 \\
      \wedge (\neg (t_2.\MatchRow(x_2) = x_2) \\
      \qquad \Rightarrow t_0.\ActionDataRow(y) \% 2 = 0)
    \end{array}
  \end{array} \\\\
  \wedge \neg (t_0.\MatchRow(x_0) = x_0)
  \Rightarrow
  \begin{array}[t]{l}
    t_1.\MatchRow(x_1) = x_1 \Rightarrow t_1.\ActionDataRow(y) \% 2 = 0 \\
    \wedge \, \neg(t_1.\MatchRow(x_1) = x_1 \Rightarrow \bot)\\
    \end{array}
\end{array}
\]
which is similar to p4v's approach.

\subsubsection{Question~\ref{question:elimination}}

Looking at these examples, its not clear that we can eliminate the dataplane
state, which occurs only in the form of match conditions. Specifically,
dataplane state occurs only in the formula as expressions the form
\(t_i.\MatchRow(x_i) = x_i \Rightarrow \varphi\), where \(\varphi\) is some
further formula that may intermingle control and dataplane state.

Inspecting the formula above, my intuition tells me that we should be able to
come up with a rewrite rule that replaces these occurences with \(\top\),
resulting in a formula with shape.

\[
\begin{array}{l}
  \begin{array}[t]{l}
    t_0.\ActionDataRow(y) \% 3 = 0 \Rightarrow
    \begin{array}[t]{l}
      t_1.\ActionDataRow(y) \% 2 = 0 \wedge t_0.\ActionDataRow(y)\% 2 = 0 \\
    \end{array} \\
    \wedge\,t_0.\ActionDataRow(y) \% 3 \neq 0 \Rightarrow
    \begin{array}[t]{l}
      t_2.\ActionDataRow(y) \% 2 = 0 \wedge t_0.\ActionDataRow(y)\% 2 = 0 \\
    \end{array}
  \end{array} \\
  \,\,\,\wedge\, t_1.\ActionDataRow(y) \% 2 = 0 \\
\end{array}
\]

which constraints only controllable variables.

However, we have completely lost the fact that no packet may miss in table
\(t_1\), which is encoded in the raw formula. More care is needed here.

\subsection{Existential Example}

For a while I've been concerned with the issue of control plane conditions of
the form \[\forall \rho_1 \in t_1, \exists \rho_2 \in t_2,\] because I was
afraid that this row existence wouldn't be captured in the predicate transformer
semantics. This is related to Question~\ref{question:rowiter}---the multiplicity
of rows is intended to enable this kind of fine-grained approach.

It seems that we \emph{can} recover the existential information from the weakest
preconditions constructed from instrumenting with only a single rule.

Consider the following program that focuses on a single header instance \texttt{h}.
\begin{figure}[H]
  \begin{minipage}{0.32\textwidth}
\begin{verbatim}
control {
  out := 0;
  h.setInvalid();
  t1.apply();
  if(h.isValid()){
    t2.apply();
    assert out!=0;
  }
}
\end{verbatim}
  \end{minipage} \begin{minipage}{0.39\textwidth}
\begin{verbatim}
table t1 {
  keys = x : exact;
  acts : set_h
  default_action : nop
}
table t2 {
  keys = h.f:exact;
  acts = {nop; set_out}
  default_action = nop;:
}
\end{verbatim}
  \end{minipage} \begin{minipage}{0.27\textwidth}
\begin{verbatim}
action set_h(f)  {
  h.setValid();
  h.f := x;
}
action set_out (o) {
  out := o
}
action nop () {}
\end{verbatim}
\end{minipage}
\end{figure}

Notice that the control plane must satisfy the following condtion for the assertion to never fail:
\[ \forall \rho_1 \in t_1.~\exists \rho_2 \in t_2.~\rho_2.\MatchRow(h.f) = \rho_1.\ActionDataRow(f) \wedge \rho_2.\ActionDataRow(o) \neq 0 \]

If we follow the single-row-per-table instrumentation strategy, we compute the
following condition:

\[\begin{array}{l}
t_1.\MatchRow(x) = x \Rightarrow \\
\quad t_2.\MatchRow(h.f) = t_1.\ActionDataRow(f) \\
\quad \wedge t_2.\ActionDataRow(o) \neq 0
\end{array}\]

which bears a striking resemblence to the condition above. We can typset them side-by-side to highlight the similarities.

\[\begin{array}{lcl}
t_1.\MatchRow(x) = x \Rightarrow && \forall \rho_1 \in t_1. \\
\quad t_2.\MatchRow(h.f) = t_1.\ActionDataRow(f) & \leadsto & \quad \exists \rho_2 \in t_2.~\rho_2.\MatchRow(h.f) = \rho_1.\ActionDataRow(f) \\
\qquad \wedge t_2.\ActionDataRow(o) \neq 0 && \qquad \wedge\rho_2.\ActionDataRow(o) \neq 0
\end{array}
\]

This provides evidence for the p4v approach, in repsonse to Question~\ref{question:rowiter}.

\subsection{Hypothesizing Transformations}

In response to Question~\ref{question:eliminations}, we can hypothesize the
following transformation rules in response to the above example, where
\(t.\mathsf{o}\) represents an arbitrary access to table \(t\), \(\texttt{d}\)
represents dataplane state,

\[\begin{array}{l>{\leadsto}cr}
t.\MatchRow(x) = x \rightarrow \phi && \forall \rho \in t. \phi \\
t.\mathsf{o} = t'.\mathsf{o'} \wedge \phi && \exists \rho \in t'. \phi \\
t.\mathsf{o} = n && \rho.\mathsf{o} = n \\
\end{array}\]

An example of each of these can be seen in the example above.

However, these rules as stated are incorrect, they must aqcuire and maintain
some context. For example, in the final rule, mapping tests liket $t.\mathsf(o)
= 0$, we need to know which \(\rho\) row-variable corresponds to the \(t\).

Not only that, but we also need to know how to bind the variables. For example,
how would we translate a rule like
\[t_1.\MatchRow = 6 \Rightarrow t_2.\MatchRow = 7\]

The set of hypothesized rules suggests that this should become
\[\forall \rho_1 \in t_1. \rho_1.\MatchRow = 6 \Rightarrow \exists \rho_2 \in t_2. \rho_2.\MatchRow = 7\]

Some random thoughts
\begin{enumerate}
\item Context seems important
\item Can we generalize this into only introducing one operator? It seems like
  both of the quantifier introduction rules could be replaced by a local
  \(\exists\)-quantifiaction,
\item We need more examples and a more formal approach. Can we derive these rules?
\end{enumerate}

\subsection{Additional Examples}
More examples are needed to investigate
Questions~\ref{question:rowiter}\&\ref{question:elimination}.

\clearpage
\section{(OUTDATED) An Incremental Problem Statement}
